Distributional Trees for Circular Data

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Adapt application to larger tree (first draft)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2420 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b78cdd4a3f32c689a6e40362553f09643946a09b

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management and renewable energy production. As a suitable
and useful application of circular regression trees, we therefore forecast the
wind direction at the airport of Innsbruck. As potential split variables we
employ several 1-hourly lagged observed weather variables (e.g., wind
direction, wind speed, relative humidity, pressure) and derived spatial and
temporal difference of these lagged observations.

The weather station of interest, Innsbruck Airport, lies at the bottom of the
Inn Valley, a deep mountain valley in the Eastern European Alps orientated
parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered in the
North with a approx. 2300\,m high mountain ridge and opens in the South into
the crest-perpendicular Wipp Valley. Therefore, mainly three different wind
regimes exist: Up-valley winds blowing from the valley mouth towards the upper
valley (from east to west), down-valley winds blowing in the direction of the
valley mouth (from west to east), and dry and warm downslope winds blowing
across the Alpine crest along the Wipp Valley towards Innsbruck (from
south-east to north-west).

As an illustration of tree based forecasts of wind direction for Innsbruck Airport, a
model based regression tree of depth two is shown in Fig.~\ref{fig:tree}.
Winds typically blow from areas of high pressure to areas of low pressure. The
first chosen split variable is therefore the pressure difference between
Innsbruck and a station located near the valley mouth. In the left branch of
the tree, with lower pressure in Innsbruck than at the valley mouth, the second
split is performed on the pressure difference between Innsbruck a station
further up the valley. This leads to the final \emph{Node~3} with low pressure
gradients within the Inn Valley and, therefore, no clear partition in the wind
direction. The final \emph{Node~4} contains measurements with a strong pressure
gradient towards the upper Inn Valley leading to up-valley winds from east to
west. The right branch of the tree is split in relative humidity as an index
for the dry downslope winds. Therefore, for a low relative humidity winds from
south-east to north-west are predominant in the final \emph{Node~6}. Whereas,
the final \emph{Node~7} contains the cases with down-valley winds towards the
valley mouth in the East.

>>>>

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management and renewable energy production. As a suitable
and useful application of circular regression trees, we therefore forecast the
wind direction at the airport of Innsbruck. Innsbruck lies at the bottom of the
Inn Valley, a deep mountain valley in the Eastern European Alps orientated
parallel to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is
bordered in the North with a approx. 2300\,m high mountain ridge and opens in
the South into the crest-perpendicular Wipp Valley. As potential split
variables we employ several 1-hourly lagged observed weather variables (e.g.,
wind direction, wind speed, relative humidity, pressure gradients along the
valley) and derived spatial and temporal difference of these lagged
observations.

As an illustration for the tree based forecast performance of predicting wind
directions at Innsbruck airport, a BIC based post-pruned regression tree of
depth three is shown in Fig.~\ref{fig:tree}. According to the fitted von Mises
distributions three distinct wind regimes: Firstly, up-valley winds blowing
from the valley mouth towards the upper valley (from east to west) as shown in
the final nodes 4 and 5. Secondly, downslope winds blowing across the Alpine
crest along the Wipp Valley towards Innsbruck (from south-east to north-west)
as predicted in nodes 7 and 8. Thirdly, down-valley winds blowing in the
direction of the valley mouth (from west to east) present in nodes 12, 14, and
15. The final node 11 with a density parameter $\kappa$ of $0.84$ captures the
cases with rather low wind speeds and therefore no clear wind direction.
Generally, the splits in the lagged wind direction perform a clustering into
the different wind regimes. As winds typically blow from areas of high pressure
to areas of low pressure and have larger wind speeds in case of stronger
pressure gradients, these respective splits in pressure gradients and wind
speed divide the predictions according to the stability of the lagged wind
direction. As an example node 3 differentiates in the pressure difference along
the valley, and the cases with a stronger pressure gradient have a higher
fitted $\kappa$ as indicator for the stability of the forecast. In other words,
in final nodes with a higher $\kappa$ most training cases are classified
correctly and therefore the density parameter is an indicator for the
uncertainty of the forecast.
